The Desert of the Reals: Floating-Point Arithmetic on Deterministic Systems

### 2024-03-09 mss.tex
Clean up with new style file.
Post near-final draft.
Post some updates

%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              Manuscript submitted for review.\\
              ~ \\
              Address author correspondence to \patp{lagrev-nocfep}.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
\fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
\fancyhead[RO]{The Desert of the Reals:  Floating-Point Arithmetic on Deterministic Systems}
\fancyfoot[LE,RO]{\thepage}

@@

\begin{abstract}
Floating-point calculations are critical to a number of special domains in modern computing, including machine learning, graphics, and scientific computing.  Numerical calculations are particularly susceptible to opaque and system-local optimizations, which can break certain guarantees for deterministic computers.  We consider the background and implementation of IEEE 754 floating-point arithmetic and options for implementing mathematics compatibly with fully reproducible and portable computing.  We consider hardware-based and software-based proposals.
\end{abstract}

@@

\begin{quote}
Floating-point operations are a technically complex subject and the extent to which developers or source code alter or test this information will depend on many factors. Apart from the general exhortation to developers to be careful and to make sure they know what they are doing, there is little of practical use that can be recommended.  \citep[p.~197]{Jones2008}
\end{quote}

@@

Early numeric computing tended to focus on problems of interest to military and national security applications, such as partial differential equations and numerical optimization.  Such calculations typically involve arrays, and linear algebra was elaborated hand-in-hand with digital computing techniques in software and hardware.  Numerics assumed prominence for a wider audience with the rise of gaming on personal computers, although these algorithms emphasized speed over exactness.\footnote{This is reflected in algorithms such as the “fast inverse square root”, which permits a degree of inaccuracy in exchange for a substantial speedup.}  To this point in the history of computing, most software either ran on a single platform for its lifetime (as with supercomputing) or did not require portably deterministic algorithms (as with gaming).\footnote{To be clear, 3D gaming algorithms are deterministic (assuming no random sources are used), but exact error values are often not reproducible across platforms, nor was this a design criterion.  The reasons for this are discussed below.}

@@

\section{The Field of the Reals}

Binary computer values are at root easily represented as non-negative integers.  However, it is frequently convenient when working with human applications to either use other numeric bases (notably decimal and hexadecimal) or to permit non-integer mathematics.

In the historical development of mathematics, logical problems in each set of numbers drove the discovery and elaboration of more elaborate algebras.  For instance, in the field of natural numbers $\mathbb{N}$, the operation of addition ($+$) or multiplication ($\times$) produces a value within the set; however, permitting subtraction ($-$) of a larger number from a smaller number can result in a value inexpressible in $\mathbb{N}$.  This motivated the introduction of the integers $\mathbb{Z}$, augmenting the numbers from zero to (positive) infinity with the negative numbers.  Division ($/$) similarly produced a crisis when applied to values which did not have a whole-number ratio between them, a situation resolved by the Pythagorean\footnote{A legendary attribution, alas, predicated on the Pythagorean discovery of the irrational numbers as a separate class thereby inducing a crisis.  https://plato.stanford.edu/entries/pythagoras/} innovation of the rational numbers or fractions as a class $\mathbb{Q}$.  Ultimately, the common reference set for engineering mathematics (and the human understanding of the continuum such as measurement) is the set of real numbers.  The set of real numbers, denoted by $\mathbb{R}$, is characterized by its continuity, implying that for any two distinct values within this set, there exists a difference, no matter how small.

Since the operations and conventions of $\mathbb{R}$ have been found to be so useful, it is desirable to extend the semantics to computer programming.  However, digital computers, by virtue of their binary representation, effectively use natural numbers $\mathbb{N}$ to represent numbers (to the limit of memory rather than positive infinity $+\infty$).  Several schemes permit a computer integer to be interpreted as if it were a number with a fractional part, including a scaling factor, fixed-point representation,\footnote{Hand-in-hand with the development of linear algebra, machines such as ENIAC and MANIAC were employed in the 1940s for solving thermonuclear reaction calculations and neutron diffusion equations.  Under the direction of John von Neumann, it appears that some calculations did experimentally involve a floating-point scheme, although this was later rejected definitively in favor of fixed-point arithmetic.  \citep[p.~3]{Kahan1997a}} and floating-point representations.

@@

\begin{lstlisting}[style=listingcode]
SEEE.EEEE.EFFF.FFFF.FFFF.FFFF.FFFF.FFFF
\end{lstlisting}

where \texttt{S} is the sign bit, \texttt{0} for positive ($+$) and \texttt{1} for negative ($-$); \texttt{E} is the exponent in base-2 (8 bits); and \texttt{F} is the significand (23 bits).  The exponent is actually calculated at an offset bias of 127 ($2^{7}$) so that a more expressive range of orders of magnitude can be covered.  The significand has an implied leading \texttt{1} bit unless all are zero.  To wit,

$$
(-1)^{\mathtt{S}} \times 2^{\mathtt{E}-127} \times 1.\mathtt{F}
$$

IEEE 754 specifies operations between numbers, including of different magnitudes.  The standard dictates behavior and provides outlines for arithmetic, but leaves algorithmic details to the implementation.  Numbers are normalized by adjusting the exponent of the smaller operand and aligning the significands, then the operations are carried out.  In practice, extended precision values are used in the intermediate steps of many algorithms, leading to greater accuracy than would otherwise be expected.\footnote{“Each of the computational operations that return a numeric result specified by this standard shall be performed as if it first produced an intermediate result correct to infinite precision and with unbounded range, and then rounded that intermediate result, if necessary, to fit in the destination’s format.”  \citep{IEEE754-2008}  Note that, per \citeauthor{Risse2016}, “there is no indication whether or not a computation with IEEE 754 is exact even if all arguments are.”  The ISO/IEC9899 C standard confesses its own fallibility:  “The floating-point model is intended to clarify the description of each floating-point characteristic and does not require the floating-point arithmetic of the implementation to be identical” \citep[fn.~21]{ISO9899}.}
% https://www.iso.org/obp/ui/#iso:std:iso-iec:9899:ed-4:v1:en

@@

\begin{enumerate}
  \item  The distance between two adjacent values changes based on the magnitude of the exponent and the distance from zero.  (The significand resolution stays the same but the exponent changes.)
  \item  There is a relative approximation error for a given bitwidth in IEEE 754, called the \emph{machine epsilon}.\footnote{Note that this is different from the smallest representable value for a given bit width; e.g., for 32-bit single-precision \texttt{float} the smallest representable value is $\mathtt{0000.0000.0000.0000.0000.0000.0000.0001} = 1\times 10^{-45}$.}
  \item  Operations between numbers of different magnitudes are particularly affected by their relative numerical horizon.
\end{enumerate}

@@

For most values of the exponent \texttt{E}, the difference between two discrete values is determined by the absolute magnitude of the significand \texttt{S}.  The difference between serial values is
\begin{align}
  \Delta\mathtt{S} &= \mathtt{000.0000.0000.0000.0000.0001}_{2} \nonumber \\
  & = 1.00000011920928955078125 - 1.0 \nonumber \\
  & = 0.00000011920928955078125 \nonumber \\
  & = 2^{-23}\textrm{.}
\end{align}

@@

However, for normalized numbers, or numbers that are left-shifted or right-shifted in order to carry out a calculation, values are determined by the \emph{relative shift} in exponent $\Delta\mathtt{E}$.  For $\mathtt{E} = 2$, for instance, the difference between serial values is $2^{-21}$.  This variable precision means that the precision of floating-point values varies across the range of representable numbers when operations take place.  Operations between two numbers of fairly different precisions are particularly vulnerable to accuracy loss, although some numerical techniques can be employed to mitigate.
% https://www.h-schmidt.net/FloatConverter/IEEE754.html

Truncation error results from terminating repeating “binaries” (by analogy with “decimals”).  Just as $\frac{1}{3} = 0.\bar{3} = 0.3333…$ has a finite precision when written in base-10, numbers that are not precise powers of two result in repeating fractions.  These necessarily terminate at the resolution of the significand.  The number and nature of truncation and rounding can significantly affect the accuracy of floating-point arithmetic and algorithms \citep{Izquierdo2006}.

@@

\begin{enumerate}
  \item  Exception handling means that the hardware may specify rounding via an overflow flag.
  \item  Extended precisions formats are not a huge deal to leave out, but extended precision arithmetic (used for intermediate results) can materially change results.
  \item  Subnormals are optional;\footnote{Subnormals are a convention that allows values smaller than the “normal” IEEE 754 smallest non-zero value.  They permit a graceflow underflow behavior, and can prevent an unintentional division by zero.} some platforms may flush them to zero or (worse) allow subnormal support to be disabled in certain cases.\footnote{“Some processors do not support subnormal numbers in hardware” \citep[p.~338]{Jones2008}.  (The risk is that this permits inadvertent division by zero.)  Various chipsets solve this exceptional behavior differently.}
\end{enumerate}

@@

For instance, IEEE 754 defines a range of numbers as “not-a-number” values, or NaNs.  Per the specification, a NaN can be a signalling NaN, meaning that it intends to flag and possibly disrupt a problematic computation;\footnote{“C support for signaling NaNs, or for auxiliary information that could be encoded in NaNs, is problematic.  Trap handling varies widely among implementations. Implementation mechanisms may trigger signaling NaNs, or fail to, in mysterious ways.  The IEC 60559 floating-point standard recommends that NaNs propagate; but it does not require this and not all implementations do.”  \citep[p.~339]{Jones2008}} or a quiet NaN, which does not raise such an exception and merely yields a result with the NaN propagated to the final result.\footnote{In Urbit, the Vere runtime unifies NaNs, meaning that any bitwise information which may be encoded in the significand field—the “NaN payload”—is thrown away.}  Not all processors implement this part of IEEE 754 correctly:  “The Motorola DSP563CCC does not support NaNs or infinities.  Floating-point arithmetic operations do not overflow to infinity; they saturate at the maximum representable value” \citep[p.~338]{Jones2008}.

As a further example, fused multiply-add (FMA) ($a \times b + c$) is implemented on certain hardware to favor \texttt{double} operations and not quadruple-precision operations \citep[p.~5]{Kahan1997}.
%\footnote{While this particular example may not strike the reader as particularly egregious, the existence of deviations from the specification raises the spectre of nondeterminism.}

@@

“Obtaining the correctly rounded result of an addition or subtraction operation requires an additional bit in the significand (as provided by the IEC 60559 guard bit) to hold the intermediate result” \citep[p.~65]{Jones2008}.

@@

Non-real arithmetic is less significant for many of the core operations of Urbit as a personal server platform.  However, gaming, machine learning, graphics, and other applications rely on floating-point calculations—preferably as fast as possible.  In fact, not only applications-oriented processes rely on determinism:  guarantees in cryptography and contractual correctness for web3; verification and validation; and code correctness analysis all require reproducible determinism.\footnote{Support for IEEE 754 is similar for support for the Markdown markup language.  Many platforms support a subset of Markdown coupled with platform-specific extensions.  (See also SQL.)  Internal references, HTML, inline \LaTeX~math mode, code block language specification, and other features see varying levels of support with GitHub, Pandoc, Obsidian, and other editors and converters.}

@@

Urbit makes much of avoiding the “ball of mud” “standard software architecture” \citep{Foote1999}.  In this design anti-pattern, a lack of guarantees and predictable behavior leads \emph{inevitably} to haphazard and illegible software bloat \citep{Foote1999}.  We can thus understand why Urbit as a platform considers even deviations in the last bit of a significand to be threads fraying the edge of sanity \citet{Monk2020a}:

@@

The situation grows more ambiguous across architectures.  \citet[p.~346]{Jones2008} presents the pathological case of a compliant platform that may use extended precision bits in the calculation of \texttt{a + b}:

\lstset{language=C}
\begin{lstlisting}
#include <stdio.h>

@@

If you controlled the compiler and runtime execution stack to a sufficient degree, could you yield deterministic floating-point arithmetic from the hardware?  “A translator that generates very high performance code is of no use if the final behavior is incorrect” \citep[p.~189]{Jones2008}; that is, optimizations often come at the cost of correctness.

@@

For instance, suppose that one intended to use the C keyword \texttt{volatile} to block certain common optimizations on a floating-point value.\footnote{In any case, this assumes a legible and enumerable set of behaviors for \texttt{volatile} which is, alas, not the case.  “\texttt{volatile} is a hint to the implementation to avoid aggressive optimization involving the object because the value of the object might be changed by means undetectable by an implementation” \citep[p.~472]{Jones2008}.  “Actions on objects so declared shall not be ‘optimized out’ by an implementation or reordered except as permitted by the rules for evaluating expressions” \citep[p.~1500]{Jones2008}.  “The \texttt{volatile} qualifier only indicates that the value of an object may change in ways unknown to the translator (therefore the quality of generated machine code may be degraded because a translator cannot make use of previous accesses to optimize the current access)” \citep[p.~963]{Jones2008}.  The same author provides examples of C code that is ambiguous in \texttt{volatile}'s semantics, pp.~1290–1291; and undefined in \texttt{volatile}'s semantics, pp.~1482–1483.}  The runtime at the level of Nock does not know if a value is considered floating-point or not.  At the level of a jet, the use of \texttt{volatile} can correctly bar certain hardware optimizations, but these need to be carefully enumerated and understood in the light of the other toolchain concerns enumerated in this section.  Strictly speaking, \texttt{volatile} only seeks to guarantee that stale calculations are not inadvertently reused due to optimization.  Without hardware optimization, the utility of an FLU for fast floating-point computations is questionable.  The risk of a jet mismatch remains high, as does a nonportable jet.\footnote{“What constitutes an access to an object that has volatile-qualified type is implementation-defined” \citep[p.~1488]{Jones2008}.  “Volatile-qualified objects can also be affected by translator optimizations” \citep[p.~1490]{Jones2008}.  The C novice may at this point wonder what the intended utility of \texttt{volatile} in fact is:  “[a] \texttt{volatile} declaration may be used to describe an object corresponding to a memory-mapped input/output port or an object accessed by an asynchronously interrupting function” \citep[p.~1499]{Jones2008}.}

@@

Finally, “[a]n implementation is not required to provide a facility for altering the modes for translation-time arithmetic, or for making exception flags from the translation available to the executing program” \citep[p.~200]{Jones2008}.  The information we purport to gain by controlling the stack in the manner above outlined is possibly not even available to the compiler and the runtime executable.

@@

\begin{quote}
A recently proposed hardware acceleration technique is to store the results of previous multiplication and division operations in a cache, reusing rather than recalculating the result whenever possible. (Dynamic profiling has found that a high percentage of these operations share the same operands as previous operations.)  \citep[p.~1148]{Jones2008}\footnote{Cf. Jones' citation of \citeauthor{Citron1998}} %no period for et al.
\end{quote}

@@

“Storing results” could also be met by the use of SPARC-style logging.  In that hardware platform, suspicious computations are flagged and hashed into a lookup table by site in the originating program.  Such events are logged not by timestamp or by computation hash but by callsite in the originating program \citep[p.~6]{Kahan1997}.\footnote{What constitutes “suspicion” is only sparsely elaborated by Kahan in that article.}  Sun implemented this in SPARC for “retrospective diagnostics” but the technique could allow a more lapidary operation for Urbit.  (Follow-on considerations include whether such computations should now be considered “bound” in a sense like that of the scry namespace.)

@@

What degree of vetting will reliably answer the gap between IEEE 754 and hardware implementation for any particular operation?  \citet[pp.~330ff.]{Jones2008} and \citet{Goldberg1991} provide a careful analysis of accuracy errors inherent to IEEE 754 as a standard, but due to the variety of possible scenarios do not treat of real compilers and chipsets much.\footnote{See particularly the note on “Common Implementations” on p.~346 of \citet{Jones2008}.}  Trivially, as demonstrated above in the Python example, $(a + b) + c \neq a + (b + c)$, and even modest reordering of operations by a zealous compiler optimization is susceptible of introducing nonportable and thus nondeterministic (in our sense) behavior.

@@

\citep{Gustafson2015}
% https://www.crcpress.com/The-End-of-Error-Unum-Computing/Gustafson/p/book/9781482239867
\citep{Gustafson2017}
% http://superfri.org/superfri/article/download/137/232
\citep{Posit2022}
% https://posithub.org/docs/posit_standard-2.pdf
\citep{Risse2016}
% https://www.vts.su.ac.rs/data/files/1/uploads/sip2016/proceedings/10_Risse_SIP2016.pdf

@@

A fixed-point representation differs from a floating-point scheme in that the exponent is fixed by the protocol or metadata and thus only the sign and significand need be included in the bit representation.  (With an offset, even the sign can be elided.)  The advantage of such a scheme is that it affords the benefits of floating-point mathematics at near-integer operation speeds (e.g. left-shift to multiply by two).  One disadvantage is that there is a smallest representable value; this lack of subnormals requires either an underflow handler or the possibility of inadvertent division by zero.  Fixed-point operations could also be used as intermediates in calculations.  (This echoes once again the idea of conversion to an intermediate representation then conversion back out to IEEE 754.)\footnote{“One solution to implementing floating-point types on processors that support fixed-point types is to
convert the source containing floating-point data operations to make calls to a fixed-point library” \citep[p.~346]{Jones2008}.  Note that the sense of our current interest is reversed.}

>>>>

\addbibresource{mss.bib}

%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              Manuscript submitted for review.\\
              ~ \\
              Address author correspondence to \patp{lagrev-nocfep}.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
\fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
\fancyhead[RO]{Floating-Point Arithmetic on Deterministic Systems}
\fancyfoot[LE,RO]{\thepage}

@@

\begin{abstract}
\noindent \sloppy
Floating-point calculations are critical to a number of special domains in modern computing, including machine learning, graphics, and scientific computing.  Numerical calculations are particularly susceptible to opaque and system-local optimizations, which can break certain guarantees for deterministic computers.  We consider the background and implementation of IEEE 754 floating-point arithmetic and options for implementing mathematics compatibly with fully reproducible and portable computing.  We consider hardware-based and software-based proposals.
\end{abstract}

@@

\begin{quote}
Floating-point operations are a technically complex subject and the extent to which developers or source code alter or test this information will depend on many factors. Apart from the general exhortation to developers to be careful and to make sure they know what they are doing, there is little of practical use that can be recommended.  \citetext{(\citet{Jones2008}, p.~197)}
\end{quote}

@@

Early numeric computing tended to focus on problems of interest to military and national security applications, such as partial differential equations and numerical optimization.  Such calculations typically involve arrays, and linear algebra was elaborated hand-in-hand with digital computing techniques in software and hardware.  Numerics assumed prominence for a wider audience with the rise of gaming on personal computers, although these algorithms emphasized speed over exactness.\footnote{This is reflected in algorithms such as the “fast inverse square root”, which permits a degree of inaccuracy in exchange for a substantial speedup.}  To this point in the history of computing, most software either ran on a single platform for its lifetime (as with supercomputing) or did not require portably deterministic algorithms (as with gaming).\footnote{To be clear, 3D gaming algorithms are deterministic (assuming no random sources are used), but exact error values are often not reproducible across platforms, nor was such portability a design criterion.  The reasons for this are discussed below.}

@@

\section{A Derivation of the Real Numbers}

\sloppy
Binary computer values are at root easily represented as non-negative integers.  However, it is frequently convenient when working with human applications to either use other numeric bases (notably decimal and hexadecimal) or to permit non-integer mathematics.

In the historical development of mathematics, logical problems in each set of numbers drove the discovery and elaboration of more elaborate algebras.  For instance, in the field of natural numbers $\mathbb{N}$, the operation of addition $+$ or multiplication $\times$ produces a value within the set; however, permitting subtraction $-$ of a larger number from a smaller number can result in a value inexpressible in $\mathbb{N}$.  This motivated the introduction of the integers $\mathbb{Z}$, augmenting the numbers from zero to (positive) infinity with the negative numbers.  Division $/$ similarly produced a crisis when applied to values which did not have a whole-number ratio between them, a situation resolved by the Pythagorean\footnote{A legendary attribution, alas, predicated on the Pythagorean discovery of the irrational numbers as a separate class thereby inducing a crisis.  https://plato.stanford.edu/entries/pythagoras/} innovation of the rational numbers or fractions as a class $\mathbb{Q}$.  Ultimately, the common reference set for engineering mathematics (and the human understanding of the continuum such as measurement) is the set of real numbers.  The set of real numbers, denoted by $\mathbb{R}$, is characterized by its continuity, implying that for any two distinct values within this set, there exists a difference, no matter how small.

\sloppy
Since the operations and conventions of $\mathbb{R}$ have been found to be so useful, it is desirable to extend the semantics to computer programming.  However, digital computers, by virtue of their binary representation, effectively use natural numbers $\mathbb{N}$ to represent numbers (to the limit of memory rather than positive infinity $+\infty$).  Several schemes permit a computer integer to be interpreted as if it were a number with a fractional part, including a scaling factor, fixed-point representation,\footnote{Hand-in-hand with the development of linear algebra, machines such as ENIAC and MANIAC were employed in the 1940s for solving thermonuclear reaction calculations and neutron diffusion equations.  Under the direction of John von Neumann, it appears that some calculations did experimentally involve a floating-point scheme, although this was later rejected definitively in favor of fixed-point arithmetic.  See \citetext{(\citet{Kahan1997a}, p.~3)} on floating-point representations.}

@@

\begin{lstlisting}[style=listingblock]
SEEE.EEEE.EFFF.FFFF.FFFF.FFFF.FFFF.FFFF
\end{lstlisting}
\noindent
where \texttt{S} is the sign bit, \texttt{0} for positive ($+$) and \texttt{1} for negative ($-$); \texttt{E} is the exponent in base-2 (8 bits); and \texttt{F} is the significand (23 bits).  The exponent is actually calculated at an offset bias of 127 ($2^{7}$) so that a more expressive range of orders of magnitude can be covered.  The significand has an implied leading \texttt{1} bit unless all are zero.  To wit,

$$
(-1)^{\texttt{S}} \times 2^{\texttt{E}-127} \times 1.\texttt{F}
$$
\noindent
IEEE 754 specifies operations between numbers, including of different magnitudes.  The standard dictates behavior and provides outlines for arithmetic, but leaves algorithmic details to the implementation.  Numbers are normalized by adjusting the exponent of the smaller operand and aligning the significands, then the operations are carried out.  In practice, extended precision values are used in the intermediate steps of many algorithms, leading to greater accuracy than would otherwise be expected.\footnote{“Each of the computational operations that return a numeric result specified by this standard shall be performed as if it first produced an intermediate result correct to infinite precision and with unbounded range, and then rounded that intermediate result, if necessary, to fit in the destination’s format.”  \citep{IEEE754-2008}  Note that, per \citeauthor{Risse2016}, “there is no indication whether or not a computation with IEEE 754 is exact even if all arguments are.”  The ISO/IEC9899 C standard confesses its own fallibility:  “The floating-point model is intended to clarify the description of each floating-point characteristic and does not require the floating-point arithmetic of the implementation to be identical” \citetext{(\citet{ISO9899}, fn.~21)}.}
% https://www.iso.org/obp/ui/#iso:std:iso-iec:9899:ed-4:v1:en

@@

\begin{enumerate}
  \item  The distance between two adjacent values changes based on the magnitude of the exponent and the distance from zero.  (The significand resolution stays the same but the exponent changes.)
  \item  There is a relative approximation error for a given bitwidth in IEEE 754, called the \emph{machine epsilon}.\footnote{Note that this is different from the smallest representable value for a given bit width; e.g., for 32-bit single-precision \texttt{float} the smallest representable value is $\texttt{0000.0000.0000.0000.0000.0000.0000.0001} = 1\times 10^{-45}$.}
  \item  Operations between numbers of different magnitudes are particularly affected by their relative numerical horizon.
\end{enumerate}

@@

For most values of the exponent \texttt{E}, the difference between two discrete values is determined by the absolute magnitude of the significand \texttt{S}.  The difference between serial values is
\begin{align}
  \Delta\texttt{S} &= \texttt{000.0000.0000.0000.0000.0001}_{2} \nonumber \\
  & = 1.00000011920928955078125 - 1.0 \nonumber \\
  & = 0.00000011920928955078125 \nonumber \\
  & = 2^{-23}\textrm{.}
\end{align}

@@

However, for normalized numbers, or numbers that are left-shifted or right-shifted in order to carry out a calculation, values are determined by the \emph{relative shift} in exponent $\Delta\texttt{E}$.  For $\texttt{E} = 2$, for instance, the difference between serial values is $2^{-21}$.  This variable precision means that the precision of floating-point values varies across the range of representable numbers when operations take place.  Operations between two numbers of fairly different precisions are particularly vulnerable to accuracy loss, although some numerical techniques can be employed to mitigate.
% https://www.h-schmidt.net/FloatConverter/IEEE754.html

Truncation error results from terminating repeating “binaries” (by analogy with “decimals”).  Just as $\frac{1}{3} = 0.\bar{3} = 0.3333\ldots{}$ has a finite precision when written in base-10, numbers that are not precise powers of two result in repeating fractions.  These necessarily terminate at the resolution of the significand.  The number and nature of truncation and rounding can significantly affect the accuracy of floating-point arithmetic and algorithms \citep{Izquierdo2006}.

@@

\begin{enumerate}
  \item  Exception handling means that the hardware may specify rounding via an overflow flag.
  \item  Extended precisions formats are not a huge deal to leave out, but extended precision arithmetic (used for intermediate results) can materially change results.
  \item  \sloppy Subnormals are optional;\footnote{Subnormals are a convention that allows values smaller than the “normal” IEEE 754 smallest non-zero value.  They permit a graceflow underflow behavior, and can prevent an unintentional division by zero.} some platforms may flush them to zero or (worse) allow subnormal support to be disabled in certain cases.\footnote{“Some processors do not support subnormal numbers in hardware” \citetext{(\citet{Jones2008}, p.~338)}.  (The risk is that this permits inadvertent division by zero.)  Various chipsets solve this exceptional behavior differently.}
\end{enumerate}

@@

For instance, IEEE 754 defines a range of numbers as “not-a-number” values, or NaNs.  Per the specification, a NaN can be a signalling NaN, meaning that it intends to flag and possibly disrupt a problematic computation;\footnote{“C support for signaling NaNs, or for auxiliary information that could be encoded in NaNs, is problematic.  Trap handling varies widely among implementations. Implementation mechanisms may trigger signaling NaNs, or fail to, in mysterious ways.  The IEC 60559 floating-point standard recommends that NaNs propagate; but it does not require this and not all implementations do.”  \citetext{(\citet{Jones2008}, p.~339)}} or a quiet NaN, which does not raise such an exception and merely yields a result with the NaN propagated to the final result.\footnote{In Urbit, the Vere runtime unifies NaNs, meaning that any bitwise information which may be encoded in the significand field—the “NaN payload”—is thrown away.}  Not all processors implement this part of IEEE 754 correctly:  “The Motorola DSP563CCC does not support NaNs or infinities.  Floating-point arithmetic operations do not overflow to infinity; they saturate at the maximum representable value” \citetext{(\citet{Jones2008}, p.~338)}.

As a further example, fused multiply-add (FMA) ($a \times b + c$) is implemented on certain hardware to favor \texttt{double} operations and not quadruple-precision operations \citetext{(\citep{Kahan1997}, p.~5)}.
%\footnote{While this particular example may not strike the reader as particularly egregious, the existence of deviations from the specification raises the spectre of nondeterminism.}

@@

“Obtaining the correctly rounded result of an addition or subtraction operation requires an additional bit in the significand (as provided by the IEC 60559 guard bit) to hold the intermediate result” \citetext{(\citet{Jones2008}, p.~65)}.

@@

Non-real arithmetic is less significant for many of the core operations of Urbit as a personal server platform.  However, gaming, machine learning, graphics, and other applications rely on floating-point calculations—preferably as fast as possible.  In fact, not only applications-oriented processes rely on determinism:  guarantees in cryptography and contractual correctness for web3; verification and validation; accounting and legal compliance; and code correctness analysis all require reproducible determinism.\footnote{Support for IEEE 754 is similar for support for the Markdown markup language.  Many platforms support a subset of Markdown coupled with platform-specific extensions.  (See also SQL.)  Internal references, HTML, inline \LaTeX~math mode, code block language specification, and other features see varying levels of support with GitHub, Pandoc, Obsidian, and other editors and converters.}

@@

Urbit makes much of avoiding the “ball of mud” “standard software architecture” \citep{Foote1999}.  In this design anti-pattern, a lack of guarantees and predictable behavior leads \emph{inevitably} to haphazard and illegible software bloat \citep{Foote1999}.  We can thus understand why Urbit as a platform considers even deviations in the last bit of a significand to be threads fraying the edge of sanity \citep{Monk2020a}:

@@

The situation grows more ambiguous across architectures.  \citetext{\citet{Jones2008}, p.~346,} presents the pathological case of a compliant platform that may use extended precision bits in the calculation of \texttt{a + b}:

\begin{lstlisting}[style=listingcode_c]
#include <stdio.h>

@@

If you controlled the compiler and runtime execution stack to a sufficient degree, could you yield deterministic floating-point arithmetic from the hardware?  “A translator that generates very high performance code is of no use if the final behavior is incorrect” \citetext{(\citet{Jones2008}, p.~189)}; that is, optimizations often come at the cost of correctness.

@@

For instance, suppose that one intended to use the C keyword \texttt{volatile} to block certain common optimizations on a floating-point value.\footnote{In any case, this assumes a legible and enumerable set of behaviors for \texttt{volatile} which is, alas, not the case.  “\texttt{volatile} is a hint to the implementation to avoid aggressive optimization involving the object because the value of the object might be changed by means undetectable by an implementation” \citetext{(\citet{Jones2008}, p.~472)}.  “Actions on objects so declared shall not be ‘optimized out’ by an implementation or reordered except as permitted by the rules for evaluating expressions” \citetext{(\citet{Jones2008}, p.~1500)}.  “The \texttt{volatile} qualifier only indicates that the value of an object may change in ways unknown to the translator (therefore the quality of generated machine code may be degraded because a translator cannot make use of previous accesses to optimize the current access)” \citetext{(\citet{Jones2008}, p.~963)}.  The same author provides examples of C code that is ambiguous in \texttt{volatile}'s semantics, pp.~1290–1291; and undefined in \texttt{volatile}'s semantics, pp.~1482–1483.}  The runtime at the level of Nock does not know if a value is considered floating-point or not.  At the level of a jet, the use of \texttt{volatile} can correctly bar certain hardware optimizations, but these need to be carefully enumerated and understood in the light of the other toolchain concerns enumerated in this section.  Strictly speaking, \texttt{volatile} only seeks to guarantee that stale calculations are not inadvertently reused due to optimization.  Without hardware optimization, the utility of an FLU for fast floating-point computations is questionable.  The risk of a jet mismatch remains high, as does a nonportable jet.\footnote{“What constitutes an access to an object that has volatile-qualified type is implementation-defined” \citetext{(\citet{Jones2008}, p.~1488)}.  “Volatile-qualified objects can also be affected by translator optimizations” \citetext{(\citet{Jones2008}, p.~1490)}.  The C novice may at this point wonder what the intended utility of \texttt{volatile} in fact is:  “[a] \texttt{volatile} declaration may be used to describe an object corresponding to a memory-mapped input/output port or an object accessed by an asynchronously interrupting function” \citetext{(\citet{Jones2008}, p.~1499)}.}

@@

Finally, “[a]n implementation is not required to provide a facility for altering the modes for translation-time arithmetic, or for making exception flags from the translation available to the executing program” \citetext{(\citet{Jones2008}, p.~200)}.  The information we purport to gain by controlling the stack in the manner above outlined is possibly not even available to the compiler and the runtime executable.

@@

\begin{quote}
A recently proposed hardware acceleration technique is to store the results of previous multiplication and division operations in a cache, reusing rather than recalculating the result whenever possible. (Dynamic profiling has found that a high percentage of these operations share the same operands as previous operations.)  \citetext{(\citet{Jones2008}, p.~1148)}\footnote{Cf. Jones' citation of \citeauthor{Citron1998}} %no period for et al.
\end{quote}

@@

“Storing results” could also be met by the use of SPARC-style logging.  In that hardware platform, suspicious computations are flagged and hashed into a lookup table by site in the originating program.  Such events are logged not by timestamp or by computation hash but by callsite in the originating program \citetext{(\citet{Kahan1997}, p.~6)}.\footnote{What constitutes “suspicion” is only sparsely elaborated by Kahan in that article.}  Sun implemented this in SPARC for “retrospective diagnostics” but the technique could allow a more lapidary operation for Urbit.  (Follow-on considerations include whether such computations should now be considered “bound” in a sense like that of the scry namespace.)

@@

What degree of vetting will reliably answer the gap between IEEE 754 and hardware implementation for any particular operation?  \citetext{\citet{Jones2008}, pp.~330ff.} and \citet{Goldberg1991} provide a careful analysis of accuracy errors inherent to IEEE 754 as a standard, but due to the variety of possible scenarios do not treat of real compilers and chipsets much.\footnote{See particularly the note on “Common Implementations” on p.~346 of \citet{Jones2008}.}  Trivially, as demonstrated above in the Python example, $(a + b) + c \neq a + (b + c)$, and even modest reordering of operations by a zealous compiler optimization is susceptible of introducing nonportable and thus nondeterministic (in our sense) behavior.

@@

TODO
\citep{Gustafson2015}
% https://www.crcpress.com/The-End-of-Error-Unum-Computing/Gustafson/p/book/9781482239867
\citep{Gustafson2017}
% http://superfri.org/superfri/article/download/137/232
\citep{Posit2022}
% https://posithub.org/docs/posit_standard-2.pdf
\citep{Risse2016}
% https://www.vts.su.ac.rs/data/files/1/uploads/sip2016/proceedings/10_Risse_SIP2016.pdf

@@

A fixed-point representation differs from a floating-point scheme in that the exponent is fixed by the protocol or metadata and thus only the sign and significand need be included in the bit representation.  (With an offset, even the sign can be elided.)  The advantage of such a scheme is that it affords the benefits of floating-point mathematics at near-integer operation speeds (e.g. left-shift to multiply by two).  One disadvantage is that there is a smallest representable value; this lack of subnormals requires either an underflow handler or the possibility of inadvertent division by zero.  Fixed-point operations could also be used as intermediates in calculations.  (This echoes once again the idea of conversion to an intermediate representation then conversion back out to IEEE 754.)\footnote{“One solution to implementing floating-point types on processors that support fixed-point types is to
convert the source containing floating-point data operations to make calls to a fixed-point library” \citetext{(\citet{Jones2008}, p.~346)}.  Note that the sense of our current interest is reversed.}
